Experience Report: Haskell in Computational Biology

### 2012-03-06 mrfy_experience_report.tex
some conclusion-ish stuff written

Haskell makes it difficult to write code in haste, because Haskell militates towards careful type-checking and clearly thought-out function contracts. Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the programmer who attempts to write code quickly with the goal of later making it work correctly. We have found MRFy to be easier to enhance and maintain than our existing C++ code bases.

>>>>

Our previous computational biology projects include an enhancement to a protein 
structural aligner that previously came from our group, Matt~\cite{matt}. 
Matt's code base contains approximately 12,000 lines of C++ code, and relies on 
data structures such as mutable oct-trees, vectors, and arrays, with 
significant amounts of clever pointer arithmetic. Our enhancement involved 
modifying Matt to incorporate sequence information along with the structural 
aligner. While we initially estimated this would be a three-month project, it 
took most of a year because the mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: since 
invariants were not clear, nearly every change resulted in new segfaults. One 
enhancement, partial alignments (essentially a cosmetic manipulation of the 
output, but requiring deep information about some of the data structures) was 
deemed infeasible. We believe we could have completed a ground-up rewrite in 
Haskell, with most of the run-time performance, in fewer than nine months.

In general, the failure modes we encounter most in C++ programming are 
segmentation faults, memory errors (due to pointer arithmetic and uninitialized 
memory) and heap exhaustion. In contrast, the failure modes we most commonly 
see in Haskell code are type-check failures, or at worst, runtime errors around 
bounds checking of arrays and vectors. While bounds checking failures have 
proved difficult to debug, they are still less so than C++ memory errors.

Haskell makes it difficult to write code in haste, because Haskell militates 
towards careful type-checking and clearly thought-out function contracts. 
Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the 
programmer who attempts to write code quickly with the goal of later making it 
work correctly. We have found MRFy to be easier to enhance and maintain than 
our existing C++ code bases. Despite the learning curve of Haskell, we 
implemented most of MRFy in roughly three months of part-time work.
